---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---

In this file we analyse some activity monitoring data. 
First, let us load it: 

## Loading and preprocessing the data

```{r Reading data, cache=TRUE}
M_data = read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"))
# M_data$start_time= with(M_data,
#                         paste(sprintf("%02d",floor(interval/60)),sprintf("%02d",interval%%60),sep = ":"))
```


## What is mean total number of steps taken per day?

```{r Histogram of steps a day}
library(dplyr)
by_date = group_by(M_data, date) %>% summarise(steps = sum(steps))
hist(by_date$steps,
     xlab = "Steps in a day", 
     main = "Histogram of steps in a day")
mean_steps_per_day = mean(by_date$steps, na.rm = TRUE)
median_steps_per_day = median(by_date$steps, na.rm = TRUE)
```

The mean number of steps per day is `r format(mean_steps_per_day, digits = 5)`, 
and the median number of steps per day is `r format(median_steps_per_day, digits = 5)`.

## What is the average daily activity pattern?

```{r Average daily activity}
by_time = group_by(M_data, interval) %>% summarise(steps = mean(steps, na.rm = TRUE))
plot(x = by_time$interval, y = by_time$steps, 
     type = "l",
     xlab = "Interval",
     ylab = "Number of steps",
     main = "Average steps at each interval")
```

## Imputing missing values

Let us now process the missing values. 
Since the only measured quantity is the number of steps, 
this is the only column where missing values are found. 
On the other hand, looking at the plot of mean steps at each hour 
seams sensible to use this data to fill the missing values, 
as there is a lot of variability between some times and others.

```{r Filling missing values}

M_data_f = M_data # M_data, filled

for (i in seq_along(M_data_f$steps)){
  if (is.na(M_data$steps[i])){
    M_data_f$steps[i] = by_time[by_time$interval == M_data_f$interval[i],]$steps
    M_data_f$Imputed[i] = TRUE
  }
  else{M_data_f[M_data_f$date == M_data[i,]$date,]$Imputed = FALSE}
}
```

Is there any missing value left? `r any(is.na(M_data_f$steps))`.

Now we redo the histogram with the imputed values: 

```{r Histogram with imputed values}
library(ggplot2)
by_date_f = group_by(M_data_f, date) %>% summarise(steps = sum(steps), Imputed = Imputed)

ggplot(by_date_f, aes(steps, fill = factor(Imputed, levels = c(TRUE, FALSE)))) + geom_histogram(bins = 5) + labs(fill="Imputed", title = "Histogram of steps in a day") + xlab("Steps in a day") + ylab("Count") + theme(plot.title = element_text(hjust = 0.5))

mean_steps_per_day_f = mean(by_date$steps, na.rm = TRUE)
median_steps_per_day_f = median(by_date$steps, na.rm = TRUE)
```

With the imputed values the mean number of steps per day is `r format(mean_steps_per_day_f, digits = 5)`, 
and the median number of steps per day is `r format(median_steps_per_day_f, digits = 5)`. 
That is, they are not affected at all. 
Furthermore, we can see in the histogram above that all the days with imputed values are in the central bar. 

## Are there differences in activity patterns between weekdays and weekends?

```{r Weekdays and weekends comparison}

M_data_f["is_weekend"] = factor(weekdays(M_data_f$date) %in% c("Friday", "Saturday", "Sunday"), 
                                labels = c("Weekdays", "Weekend"))

by_time_f = group_by(M_data_f, interval, is_weekend) %>% summarise(steps = mean(steps, na.rm = TRUE))

ggplot(by_time_f, aes(x = interval, y = steps), type = "l") + geom_line() + facet_grid(rows = vars(is_weekend)) + labs(xlab = "Interval", ylab = "Steps", title = "Average number of steps by type of day") + theme(plot.title = element_text(hjust = 0.5))
```